[Re] Chaos in a long-term experiment with a plankton community

### 2016-02-12 article/article.tex
latest build of pdf

The data are available as an Excel file supplement to
\href{http://onlinelibrary.wiley.com/doi/10.1111/j.1461-0248.2009.01391.x/abstract}{an
Ecology Letters publication} \textcite{Beninca2009}. The Excel file
contains several datasheets. Two were particularly important, as they
were the source of the raw data (one contains original species
abundances, the other the nutrient concentrations). We saved these two
datasheets, without any alteration, as comma separated value (csv) text
files. In the code associated with this reproduction, these data files
are read from the associated github repository.

Another datasheet (we also saved this as csv file, in order to use it in
this reproduction). We also received a dataset direct from Steve Ellner,
see below for details.

The original species abundance data contained errors (i.e., a few
numeric values had a comma in place of a period as the decimal
separator, and a duplicated date).

@@

The R language and environment for statistical computing and graphics
was used to make the reproduction. Additional R packages required are
specified in the code associated with this reproduction.

The code for making the analyses and figures presented in this
reproduction resides in an R markdown document, as well as a source file
containing some required functions. Some of the code takes several
minutes to run, so an intermediate data file is provided with results
from this code.

@@

Highlighted in the text of the original paper were: negative
correlations of picophytoplankton with protozoa, and of
nanophytoplankton both with rotifers and calanoid copepods, positive
correlation of picophytoplankton with calanoid copepods, negative
correlation between bacteria and ostracods, and positive correlation
between bacteria and phosphorus. All of these correlations were
qualitatively reproduced.

@@

Spectral analyses in the original paper were presented graphically in
figures S3 (raw spectrograms) and S4 (Welch periodograms). These graphs
were, apparently, visually inspected to reveal: ``fluctuations covered a
range of different periodicities'', and ``picophytoplankton, rotifers
and calanoid copepods seemed to fluctuate predominantly with a
periodicity of about 30 days.'' It is unclear how these conclusions were
derived from figures S3 and S4 of the original article. Our reproduced
spectra (not shown here, but code provided) were a reasonably close
match to the spectra in the original article.

@@

Reproduced divergence rates (figure \ref{fig:divergence}) and Lyapunov
exponents (figure \ref{fig:LE_comparison}) were somewhat different from
those in the original article.. The original article states: ``the
distance between initially nearby trajectories increased over time, and
reached a plateau after about 20--30 days''. The reproduced results
appear not inconsistent with this statement, except for one group of
species (Harpacticoids). The original article also stated that the
analyses ``yielded significantly positive Lyapunov exponents of
strikingly similar value for all species (Fig. 3; mean exponent = 0.057
per day, s.d. = 0.005 per day, n = 9)''. Reproduced exponents had very
similar mean value, but had much larger standard deviation (mean = 0.060
and s.d. = 0.014).

@@

The original paper reported global Lyapunov exponent calculated via two
modelling approaches (neural network and generalised additive models
{[}GAMs{]}). Only the GAM approach was reproduced, with the assistance
of code donated by Stephen Ellner. The original article obtained a
global Lyapunov exponent of λ=0.08 day-1. The reproduced value was 0.04.
We did not reproduce the bootstrapping used to give confidence intervals
around this estimate.

@@

The reproduced results were qualitatively identical to those in the
original article. For example, all Lyapunov exponents estimated by
direct method were positive, as in the original article, consistent with
chaotic dynamics. Quantitative differences may have resulted from
difference in algorithms used. For example,the original used the
\href{http://www.mpipks-dresden.mpg.de/~tisean/}{Tisean software} to
calculate Lyapunov exponents. As this was available from CRAN
\href{http://cran.r-project.org/web/packages/RTisean/index.html}{until
mid 2014} and since it is a bit less well integrated with R, we instead
use the tseriesChaos package \textcite{tseriesChaos}, which in any case
was largely inspired by the TISEAN project. In addition, there may have
been some difference in algorithm parameters, as not all parameters
required by the functions we used were reported in the original ms.

>>>>

The data are available as an Excel file supplement to
\href{http://onlinelibrary.wiley.com/doi/10.1111/j.1461-0248.2009.01391.x/abstract}{an
Ecology Letters publication} \textcite{Beninca2009}. The Excel file
contains several datasheets. Two were particularly important, as they
were the source of the raw data (one contains original species
abundances, the other the nutrient concentrations). We saved these two
datasheets, without any alteration, as comma separated value (csv) text
files. In the code associated with this reproduction, these data files
are read from the associated github repository. Another datasheet
contained transformed data (we also saved this as csv file, in order to
use it in this reproduction). We also received a dataset direct from
Steve Ellner, see below for details.

@@

The R language and environment for statistical computing and graphics
was used to make the reproduction. Additional R packages required are
specified in the code associated with this reproduction. The code for
making the analyses and figures presented in this reproduction resides
in an R markdown document, as well as a source file containing some
required functions. Some of the code takes several minutes to run, so an
intermediate data file is provided with results from this code.

@@

Highlighted in the text of the original paper were: negative
correlations of picophytoplankton with protozoa, and of
nanophytoplankton both with rotifers and calanoid copepods, positive
correlation of picophytoplankton with calanoid copepods, negative
correlation between bacteria and ostracods, and positive correlation
between bacteria and phosphorus. All of these correlations were at least
qualitatively reproduced.

@@

Spectral analyses in the original paper were presented graphically in
figures S3 (raw spectrograms) and S4 (Welch periodograms). The original
article states: ``fluctuations covered a range of different
periodicities'', and ``picophytoplankton, rotifers and calanoid copepods
seemed to fluctuate predominantly with a periodicity of about 30 days.''
Our reproduced spectra (not shown here, but code provided) were a
reasonably close match to the spectra in the original article.

@@

Reproduced divergence rates (figure \ref{fig:divergence}) and therefore
Lyapunov exponents (figure \ref{fig:LE_comparison}) were somewhat
different from those in the original article. The original article
states: ``the distance between initially nearby trajectories increased
over time, and reached a plateau after about 20--30 days''. The
reproduced results are consistent with this statement. The original
article also stated that the analyses ``yielded significantly positive
Lyapunov exponents of strikingly similar value for all species (Fig. 3;
mean exponent = 0.057 per day, s.d. = 0.005 per day, n = 9)''.
Reproduced exponents had very similar mean value, but had larger
standard deviation (mean = 0.060 and s.d. = 0.014).

@@

The original paper reported a global Lyapunov exponent calculated via
two modelling approaches (neural network and generalised additive models
{[}GAMs{]}). Only the GAM approach was reproduced, with the assistance
of code donated by Stephen Ellner. The original article obtained a
global Lyapunov exponent of λ=0.08 day-1. The reproduced value was 0.04.
We did not reproduce the bootstrapping used to give confidence intervals
around this estimate.

@@

The reproduced results were qualitatively identical to those in the
original article, and therefore support the conclusion of chaotic
dynamics produced by non-linear species interactions. For example, all
Lyapunov exponents estimated by direct method were positive, as in the
original article, consistent with chaotic dynamics. Quantitative
differences may have resulted from difference in algorithms used. For
example,the original used the
\href{http://www.mpipks-dresden.mpg.de/~tisean/}{Tisean software} to
calculate Lyapunov exponents. As this was available from CRAN
\href{http://cran.r-project.org/web/packages/RTisean/index.html}{until
mid 2014} and since it is a bit less well integrated with R, we instead
use the tseriesChaos package \textcite{tseriesChaos}, which in any case
was largely inspired by the TISEAN project. In addition, there may have
been some difference in algorithm parameters, as not all parameters
required by the functions we used were reported in the original ms.
